Formula translation in \changed{Blitz++, NumPy} and modern Fortran: a~case~study~of~the~language~choice tradeoffs

### 2014-02-20 paper/doc/paper.tex
changes before sending LaTeX files to Sci Prog

%%%%% <arXiv>
%\documentclass[final,5p,times,twocolumn]{elsarticle}
%%%%% </arXiv>

%%%%% <sciprog>
\documentclass[twocolumn]{article}
\usepackage{lineno,authblk,latexsym,amsmath,float}
\usepackage[colorlinks=true]{hyperref}
\usepackage[hmargin=.55in,vmargin=.75in]{geometry}
\usepackage[numbers]{natbib}
\setlength{\columnsep}{1cm}
%%%%% </sciprog>

% highlighting (or not) the modified portions of the text
\newcommand{\changed}[1]{{\color{blue}#1}}
%\newcommand{\changed}[1]{#1}

\usepackage{soul}
\newcommand{\removed}[1]{{\color{blue}\st{#1}}}

@@

% fr compact section headers
\usepackage[compact]{titlesec}
\titlespacing*{\section}{0pt}{*0}{0pt}

% for source code listings
\usepackage{fancyvrb}
\fvset{
  frame=single,
  fontsize=\relscale{.6},
  fontfamily=courier,
  framerule=.2mm,
  framesep=.7mm,
  numbers=left,
  xleftmargin=4mm,
  numbersep=.25mm,
  firstnumber=last
}
\input{pygments}

@@

\newcommand{\codecpp}[4]{%
  \linespread{.8}
  \addtocounter{lstnocpp}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c+c1}{//#2}}%
  \renewcommand*\FancyVerbStopString{\PY{c+c1}{//#3}}%
  \setcounter{FancyVerbLine}{\thelinenocpp}%
  \fvset{label={listing~C.\thelstnocpp~(C++)},rulecolor=\color{black},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenocpp}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}
\newcommand{\codepyt}[4]{%
  \linespread{.8}
  \addtocounter{lstnopyt}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{\PYZsh{}#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{\PYZsh{}#3}}
  \setcounter{FancyVerbLine}{\thelinenopyt}%
  \fvset{label={listing~P.\thelstnopyt~(Python)},rulecolor=\color{blue},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenopyt}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}
\newcommand{\codefor}[4]{%
  \linespread{.8}
  \addtocounter{lstnofor}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{!#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{!#3}}
  \setcounter{FancyVerbLine}{\thelinenofor}%
  \fvset{label={listing~F.\thelstnofor~(Fortran)},rulecolor=\color{red},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenofor}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}

@@

\linespread{2}

\begin{document}
    \linenumbers

@@

    \author{~}
%    \author[1]{Sylwester Arabas\footnote{Correspondence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- 0048225546827 -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
%    \author[1]{Dorota Jarecka}
%    \author[1]{Anna Jaruga}
%    \author[2]{Maciej Fijałkowski}

%    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
%    \affil[2]{PyPy Team}

    \onecolumn
    \maketitle

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers constitute implementations of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes serve as examples for how the application of 
          object-oriented programming (OOP) techniques and new language constructs from
          C++11 and Fortran 2008 allow to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        All performance tests are carried out using free and open-source compilers.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license.

        ~\\
        {\bf Keywords:} object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
    \end{abstract}

  \twocolumn

  \tableofcontents

@@

  \vspace{-5ex}
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of~the~$\psi$~and~the~Courant 
    number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

  \vspace{-5ex}
  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

@@

  \vspace{-5ex}
  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:

  \vspace{-5ex}
  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}
  \vspace{-3.5ex}
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  \vspace{-3.5ex}
  and the actual two-dimensional donor-cell formula:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}
  \vspace{-3.5ex}
  The same formul\ae~are expressed as follows in Python:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  \codepyt{code-pyt-listings.py}{listing08}{listing09}{1}
  \vspace{-3.5ex}
  The Fortran counterparts are: 
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  \codefor{code-for-listings.f}{listing07}{listing08}{1}
  \codefor{code-for-listings.f}{listing08}{listing09}{1}
  \codefor{code-for-listings.f}{listing09}{listing10}{1}

@@

  \vspace{-5ex}
  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:

  \vspace{-5ex}
  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:

  \vspace{-5ex}
  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

@@

%%%%%% <arXiv>
%  \clearpage
%%%%%% </arXiv>
  \appendix

@@

  \begin{figure}[bh!]
    \caption{\label{fig:mpdata}
      Plot generated by the program given in listing~C.20.
      The top panel shows initial signal shape (at time t=0).
      The subsequent panels show snapshots of the advected field after 75 timesteps
        from three different simulations: donorcell (or 1 MPDATA iteration), MPDATA
        with two iterations and MPDATA with 44 iterations.
      The colour scale and the wire-frame surface correspond to signal amplitude.
      See appendix~\ref{app:usage} for discussion.
    }
  \end{figure}
  \includegraphics[height=.7\textheight]{demo}

>>>>

\documentclass[twocolumn]{article}
\usepackage{authblk,latexsym,amsmath,float}
\usepackage[runin]{abstract}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage[hmargin=.55in,vmargin=.75in]{geometry}
\usepackage[numbers]{natbib}

% do-nothing commands
\newcommand{\changed}[1]{#1}
\newcommand{\removed}[1]{#1}

@@

% for compact section headers
\usepackage[compact]{titlesec}

% for source code listings
\usepackage{fancyvrb}
\fvset{
  frame=single,
  fontsize=\relscale{.66},
  fontfamily=courier,
  framerule=.2mm,
  framesep=.7mm,
  numbers=left,
  xleftmargin=4mm,
  numbersep=.25mm,
  firstnumber=last
}
\input{pygments}

@@

\newcommand{\codecpp}[4]{%
  \addtocounter{lstnocpp}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c+c1}{//#2}}%
  \renewcommand*\FancyVerbStopString{\PY{c+c1}{//#3}}%
  \setcounter{FancyVerbLine}{\thelinenocpp}%
  \fvset{label={listing~C.\thelstnocpp~(C++)},rulecolor=\color{black},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenocpp}{\value{FancyVerbLine}}%
}
\newcommand{\codepyt}[4]{%
  \addtocounter{lstnopyt}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{\PYZsh{}#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{\PYZsh{}#3}}
  \setcounter{FancyVerbLine}{\thelinenopyt}%
  \fvset{label={listing~P.\thelstnopyt~(Python)},rulecolor=\color{blue},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenopyt}{\value{FancyVerbLine}}%
}
\newcommand{\codefor}[4]{%
  \addtocounter{lstnofor}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{!#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{!#3}}
  \setcounter{FancyVerbLine}{\thelinenofor}%
  \fvset{label={listing~F.\thelstnofor~(Fortran)},rulecolor=\color{red},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenofor}{\value{FancyVerbLine}}%
}

@@

\begin{document}

@@

    \date{}

    \author[1]{Sylwester Arabas}%\footnote{Correspondence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
    \author[1]{Dorota Jarecka}
    \author[1]{Anna Jaruga}
    \author[2]{Maciej Fijałkowski}

    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
    \affil[2]{PyPy Team}

    \twocolumn[
      \maketitle

      \begin{onecolabstract}
	  Three object-oriented implementations of a prototype solver of the advection equation are introduced.
	  The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
	  The solvers constitute implementations of the Multidimensional Positive-Definite 
	    Advective Transport Algorithm (MPDATA).
	  The introduced codes serve as examples for how the application of 
	    object-oriented programming (OOP) techniques and new language constructs from
	    C++11 and Fortran 2008 allow to reproduce the mathematical notation 
	    used in the literature within the program code.
	  A discussion on the tradeoffs of the programming language choice is presented.
	  The main angles of comparison are code brevity and syntax clarity
	    (and hence maintainability and auditability) as well as performance.
	  All performance tests are carried out using free and open-source compilers.
	  In the case of Python, a significant performance gain is observed when switching from the standard 
	    interpreter (CPython) to the PyPy implementation of Python.
	  Entire source code of all three implementations is embedded in the text and is licensed
	    under the terms of the GNU GPL license.

	  ~\\
	  {\bf Keywords:} object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
	  ~\\
      \end{onecolabstract}
  ]

  \tableofcontents
  \newpage

@@

  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of~the~$\psi$~and~the~Courant 
    number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

@@

  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:

  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  and the actual two-dimensional donor-cell formula:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}
  The same formul\ae~are expressed as follows in Python:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  \codepyt{code-pyt-listings.py}{listing08}{listing09}{1}
  The Fortran counterparts are: 
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  \codefor{code-for-listings.f}{listing07}{listing08}{1}
  \codefor{code-for-listings.f}{listing08}{listing09}{1}
  \codefor{code-for-listings.f}{listing09}{listing10}{1}

@@

  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:

  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:

  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

@@

  \newpage
  \appendix

@@

  \begin{figure}[bh!]
    \caption{\label{fig:mpdata}
      Plot generated by the program given in listing~C.20.
      The top panel shows initial signal shape (at time t=0).
      The subsequent panels show snapshots of the advected field after 75 timesteps
        from three different simulations: donorcell (or 1 MPDATA iteration), MPDATA
        with two iterations and MPDATA with 44 iterations.
      The colour scale and the wire-frame surface correspond to signal amplitude.
      See appendix~\ref{app:usage} for discussion.
    }
  \end{figure}
  \begin{center}
    \includegraphics[height=.83\textheight]{demo}
  \end{center}
